---
title: "practical-machine-learning-project"
author: "Ahmed Bahloul"
date: "11/11/2020"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Background

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: <http://groupware.les.inf.puc-rio.br/har> (see the section on the Weight Lifting Exercise Dataset).

**Note**: The data for this project come from this source: <http://groupware.les.inf.puc-rio.br/har> , Reference: Velloso, E.; Bulling, A.; Gellersen, H.; Ugulino, W.; Fuks, H. Qualitative Activity Recognition of Weight Lifting Exercises. Proceedings of 4th International Conference in Cooperation with SIGCHI (Augmented Human '13) . Stuttgart, Germany: ACM SIGCHI, 2013.

Read more: <http://groupware.les.inf.puc-rio.br/har#ixzz6dTnNnnRf>


The goal of this project is to predict the manner in which they did the exercise. This is the **"classe"** variable in the training set. We will create a report describing how we built our model, how we used cross validation, what we think the expected out of sample error is, and why we made the choices for the model selection,  we will also use our prediction model to predict 20 different test cases.

## Importing Libraries  
```{r echo = TRUE}
library(caret)
library(ggplot2)
library(lattice)
library(parallel)
library(doParallel)

```

## Loading Data 

```{r echo = TRUE}
training <- read.csv("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv") 
validation <-  read.csv("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv")
dim(training); dim(validation)
```

We see that the training data has `r dim(training)[1]` observations,  and `r dim(training)[2]` variables which need to be explored and cleaned if they have any issues like missing values. 

## Exploratory data analysis 

Before we start our prediction analysis, we need to cleanse our dataset by removing missing values and not related variable to our analysis. we will do the following cleansing: 
* Removing Missing Values Columns 
* Removing Zero Values Columns
* Removing not related Variables by name 
* Removing Near Zero Predictors 
We should also repeat every cleansing process to the final validation dataset.  

```{r echo = TRUE}
training <- training[,colMeans(is.na(training)) < .9]
training <- training[,colSums(is.na(training)) ==0]
#validation <- validation[,colMeans(is.na(validation)) < .9]
#validation <- validation[,colSums(is.na(validation)) ==0]
dim(training); dim(validation)
Col_names_remove <- c('user_name','raw_timestamp_part_1', 'raw_timestamp_part_2', 'cvtd_timestamp', 'new_window', 'num_window', 'X')
training <- training[,!(names(training) %in% Col_names_remove)]
#validation <- validation[,!(names(validation) %in% Col_names_remove)]
dim(training); dim(validation)
Near_Zero_Variance <- nearZeroVar(training)
dim(training); dim(validation)
training <- training[, -Near_Zero_Variance]
#validation <- validation[,-Near_Zero_Variance]
dim(training); dim(validation)
```

Now Our training dataset is ready to proceed with prediction, but we need to check the Multicollinearity between the predictors and remove the highly correlated variables. 


``` {r echo= TRUE}
corr_vars <- cor(training[, -53])
high_corr_vars <- findCorrelation(corr_vars, cutoff = 0.75)
training <- training[, -high_corr_vars]
```


## Preparing dataset for prediction

We will split the training to 70% as train data and 30% as test data to use the cross validation in order to validate our fitted model. 

```{r echo = TRUE}
train_in <- createDataPartition(training$classe ,  p = 0.7, list = FALSE)
trainData <- training[train_in,]
testData <- training[-train_in,]
```

Now we have `r dim(trainData)[1]` observations and `r dim(trainData)[2]` Variables in the training data set,  we will use this to train our prediction models, we will try the following machine learning algorithms: 
1.    Random Forest
2.    Gradient Boosted Trees
3.    Support Vector Machine 


### 1.    Random Forest

```{r echo = TRUE}
set.seed(1254)
cluster <- makeCluster(detectCores() - 1)
registerDoParallel(cluster)
control <- trainControl(method="cv", number=3, allowParallel = TRUE )
system.time(rf_Model <- train(classe ~ .,
               data = trainData,
               method = "rf",
               trControl=control))
stopCluster(cluster)
registerDoSEQ()
rf_Predict <- predict(rf_Model, testData)
rf_cm <- confusionMatrix(rf_Predict,factor(testData$classe))
rf_cm
```

### 2. Gradient Boosted Trees 

``` {r echo = TRUE}
#control_gbm <- trainControl(method="cv", number=3, verboseIter=F)
#gbm_Model <- train(classe ~. , data=trainData , method = "gbm" , trControl = control_gbm , tuneLength = 5, verbose = F)
#gbm_predict <- predict(gbm_Model, testData)
#gbm_cm <- confusionMatrix(gbm_Predict,factor(testData$classe))
#gbm_cm
```

### 3.Support Vector Machine 

``` {r echo = TRUE}
control_svm <- trainControl(method="cv", number=3, verboseIter=F)
svm_Model <- train(classe ~. , data=trainData , method = "svmLinear" , trControl = control_svm , tuneLength = 5, verbose = F)
svm_predict <- predict(svm_Model, testData)
svm_cm <- confusionMatrix(svm_predict,factor(testData$classe))
svm_cm
```

## Model Selection & Predictions on Validation data 

As we see both Random Forest & Gradient boosted Model has a very good Accuracy and a little out of sample errors rather than support vector machine so I am going to select *Random Forest* to predict the 20 class in the Validation data set. 

``` {r echo = TRUE}
predict20 <- predict(rf_Model,newdata=validation) 
print(predict20)
```
